---
title: "Gapminder Report"
author: "Joel Nitta"
date: "5/11/2022"
output: word_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
library(tidyverse)
library(rmarkdown)
```

## Data

```{r load-data}
# Load raw gapminder data
gapminder_raw <- read_tsv("data/gapminder.tsv")
# Create column with relative year starting at 0 for model
gapminder <- mutate(gapminder_raw, year_rel = year - min(year))
```

The `gapminder` dataset includes `r gapminder$country %>% n_distinct()` countries with per capita GDP, population, and life expectancy data from `r gapminder$year %>% min()` to `r gapminder$year %>% max()`.

## Methods

```{r model-function}
# Define function for fitting model

#' Fit a model of life expectancy by year
#'
#' @param data Tibble; Input gapminder data for one country. Must include columns
#' 'lifeExp', 'year_rel', and 'country'
#'
#' @return Tibble; results of fitting a linear model of life expectancy by year
#' 
fit_life_exp_mod <- function(gapminder) {
  lm(lifeExp ~ year_rel, gapminder) %>%
    broom::tidy() %>%
    filter(term != "(Intercept)")
}
```

```{r fit-model}
# Loop over data and fit model
life_exp_mods <-
  gapminder %>%
	nest(data = -country) %>%
	mutate(mod_res = map(data, fit_life_exp_mod)) %>%
	select(-data) %>%
	unnest(cols = mod_res)
```

We fit a linear model of life expectancy by year for each country.

## Results

Asian countries showed the biggest gains in life expectancy over time, with European countries the least.
African and American countries have a wide spread of values.

```{r plot-mod, echo = FALSE, include = TRUE, fig.cap= "Results of modeling life expectancy over time by country. Only significant models shown (*p* < 0.05)."}
# Tidy modeling output data
life_exp_mods_for_plot <-
life_exp_mods %>%
  filter(term == "year_rel", p.value < 0.05) %>%
  left_join(
    unique(select(gapminder, country, continent)), 
    by = "country"
  ) %>%
  mutate(country = fct_reorder(country, estimate))

# Plot the modeling results
ggplot(
  life_exp_mods_for_plot,
  aes(x = estimate, y = country, fill = continent)) +
  geom_errorbarh(
    aes(xmax = estimate + std.error, xmin = estimate - std.error),
    color = "grey40") +
  geom_col() +
  scale_x_continuous(name = "Change in life expectancy per year") +
  scale_fill_brewer(palette = "Set1", name = "Continent") +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )
```
